Reducing Linearizability of Priority Queues and More Data Structures to State Reachability

### 2017-07-12 final/co-regular-EPQ.tex
decidability

\begin{figure}[t]
  \centering
  %\includegraphics[width=0.8 \textwidth]{figures/PIC-Enumeration-WitnessAutomata.pdf}
  \includegraphics[width=0.8 \textwidth]{PIC-WitnessAutomata-For1.pdf}
%\vspace{-10pt}
  \caption{A register automaton capturing the scenario in Figure~\ref{fig:five enumerations}(a). We use the following notations: $A_1 = A \cup \{ \textit{call}(\textit{put},c,<r)\}$, $A_2 = A_1 \cup \{ \textit{ret}(\textit{put},b,=r) \}$, $A_3 = A_2 \cup \{ \textit{ret}(\textit{rm},c) \}$, $A_4 = A \cup \{ \textit{ret}(\textit{put},b,=r), \textit{ret}(\textit{rm},c) \}$, where $A = \{ \textit{call}(\textit{put},\top,\textit{true})$,$\textit{ret}(\textit{put},\top,\textit{true})$,$\textit{call}(\textit{rm},\top)$,$\textit{ret}(\textit{rm},\top)$,$\textit{call}(\textit{rm},\textit{empty})$,$\textit{ret}(\textit{rm},\textit{empty})\}$.}
  \label{fig:an enumeration and its witness automaton}
\end{figure}

@@

We describe a class $\mathcal{C}$ of data-independent implementations for which linearizability w.r.t. $\seqPQ$ is decidable. The implementations in $\mathcal{C}$ allow an unbounded number of values but a bounded number of priorities. Each method manipulates a finite number of local variables which store Boolean values, or data values from $\mathbb{D}$. Methods communicate through a finite number of shared variables that also store Boolean values, or data values from $\mathbb{D}$. Data values may be assigned, but never used in program predicates (e.g., in the conditions of if-then-else and while statements) so as to ensure data independence. This class captures typical implementations, or finite-state abstractions thereof, e.g., obtained via predicate abstraction. {\color{blue}Since the $\Gamma$-complete automata $A(\Gamma)$ we define use}{\color {red}Since the $\Gamma$-complete automata $A(\Gamma)$ uses} a fixed set {\color {red}$D=\{a,b,a_1,d,e,\top\}$ of values}, we have that {\color {red}$\mathcal{C}\cap A(\Gamma)\neq\emptyset$ for some $\Gamma$ iff $\mathcal{C}_D\cap A(\Gamma)\neq\emptyset$ where $\mathcal{C}_D$ is the subset of $\mathcal{C}$ that uses only values in $D$.} {\color{blue}The set of executions $\mathcal{C}_D$ can be represented by a Petri Net since bounding the set of values allows us to represent each thread with a finite-state machine. For a fixed set of priorities $P$, the register automata $A(\Gamma)$ can be transformed to finite-state automata since the number of possible valuations of the registers is bounded.}

{\color{red} The set of executions $\mathcal{C}_D$ can be represented by a Vector Addition Systems with States (VASS), since both values and priorities are finite, which implies that each thread and register automata can be transformed into finite-state automata. To obtain this transformation, the states of the VASS represent the global variables of $\mathcal{C}_D$, while each counter of the VASS then represents the number of threads which are at a particular control location within a method, with a certain valuation of the local variables. In this way we transform linearizability problem of priority queue into state-reachability problem of VASS, which is EXPSPACE-complete. When we consider only finite number of threads, the value of counters are bounded and then this problem is PSPACE \cite{DBLP:conf/ac/Esparza96}.

On the other hand, we can mimic transitions of VASS by a priority queue implementation. To obtain this, we keep $\textit{rm}$ unchanged, while in each $\textit{put}$ method, we first put a value into priority queue, and then either simulate one step of transitions or record one increase/decrease of a counter, similarly as that in \cite{conf/esop/BouajjaniEEH13}. In this way we transform the state-reachability problem of VASS into linearizability problem of priority queue. Based on above discussion, we have the following complexity result.} %The detailed proof can be found in Appendix \ref{subsec:appendix proof and definition in subsection decidability result}.

%\begin{restatable}{theorem}{ComplexityOfPriorityQueue}
%\label{theorem:complexity of priority queue}
%Verifying whether an implementation in $\mathcal{C}$ is linearizable w.r.t. $\seqPQ$ is PSPACE-complete for a fixed number of %threads, and EXPSPACE-complete otherwise.
%\end{theorem}

>>>>

\begin{figure}[t]
  \centering
  %\includegraphics[width=0.8 \textwidth]{figures/PIC-Enumeration-WitnessAutomata.pdf}
  \includegraphics[width=0.7 \textwidth]{PIC-WitnessAutomata-For1.pdf}
%\vspace{-10pt}
  \caption{A register automaton for the case in Figure~\ref{fig:five enumerations}(a), where $A_1 = A \cup \{ \textit{call}(\textit{put},c,<r)\}$, $A_2 = A_1 \cup \{ \textit{ret}(\textit{put},b,=r) \}$, $A_3 = A_2 \cup \{ \textit{ret}(\textit{rm},c) \}$, $A_4 = A \cup \{ \textit{ret}(\textit{put},b,=r), \textit{ret}(\textit{rm},c) \}$, where $A = \{ \textit{call}(\textit{put},\top,\textit{true})$,$\textit{ret}(\textit{put},\top,\textit{true})$,$\textit{call}(\textit{rm},\top)$,$\textit{ret}(\textit{rm},\top)$,$\textit{call}(\textit{rm},\textit{empty})$,$\textit{ret}(\textit{rm},\textit{empty})\}$.}
  \label{fig:an enumeration and its witness automaton}
\end{figure}

@@

We describe a class $\mathcal{C}$ of data-independent implementations for which linearizability w.r.t. $\seqPQ$ is decidable. The implementations in $\mathcal{C}$ allow an unbounded number of values but a bounded number of priorities. Each method has a finite set of local variables storing Boolean values or values from $\mathbb{D}$. Methods communicate through a finite number of shared variables interpreted also as Booleans or values from $\mathbb{D}$. To ensure data independence, values in $\mathbb{D}$ may be assigned, but never used in Boolean expressions (e.g., of if-then-else statements). This class captures typical implementations, or finite-state abstractions thereof, e.g., obtained via predicate abstraction. The $\Gamma$-complete automata we define use a fixed set $D=\{a,b,c,a_1,\top\}$ of values ($a_1$ is needed to deal with the second item in Lemma~\ref{lemma:ob order has bounded length}). Therefore, for any $\Gamma$, $\mathcal{C}\cap A(\Gamma)\neq\emptyset$ iff $\mathcal{C}_D\cap A(\Gamma)\neq\emptyset$, where $\mathcal{C}_D$ is the subset of $\mathcal{C}$ that uses only values in $D$.

The set of executions $\mathcal{C}_D$ can be represented by a Vector Addition System with States (VASS). Since the set of values and priorities is bounded, each method invocation can be represented by a finite-state automaton (see~\cite{conf/esop/BouajjaniEEH13}). For a fixed set of priorities $P\subseteq \mathbb{P}$, the register automata $A(\Gamma)$ can be transformed to finite-state automata (the number of valuations of the registers is bounded). Thus, checking linearizability of an implementation in $\mathcal{C}$ is PSPACE when the number of threads is bounded, and EXPSPACE, otherwise. Moreover, reachability in VASSs can be reduced to checking linearizability of such an implementation. Essentially, given an instance of the VASS reachability problem, one can define a priority queue implementation where the $\textit{put}$ methods behave correctly and additionally, they include the code of the VASS simulation defined in~\cite{conf/esop/BouajjaniEEH13}, and the $\textit{rm}$ methods behave correctly, except for the moment where the target state is reached, in which case they trigger a linearizability violation by returning an arbitrary value.

%{\color{red} The set of executions $\mathcal{C}_D$ can be represented by a Vector Addition Systems with States (VASS), since both values and priorities are finite, which implies that each thread and register automata can be transformed into finite-state automata. To obtain this transformation, the states of the VASS represent the global variables of $\mathcal{C}_D$, while each counter of the VASS then represents the number of threads which are at a particular control location within a method, with a certain valuation of the local variables. In this way we transform linearizability problem of priority queue into state-reachability problem of VASS, which is EXPSPACE-complete. When we consider only finite number of threads, the value of counters are bounded and then this problem is PSPACE \cite{DBLP:conf/ac/Esparza96}.
%
%On the other hand, we can mimic transitions of VASS by a priority queue implementation. To obtain this, we keep $\textit{rm}$ unchanged, while in each $\textit{put}$ method, we first put a value into priority queue, and then either simulate one step of transitions or record one increase/decrease of a counter, similarly as that in \cite{conf/esop/BouajjaniEEH13}. In this way we transform the state-reachability problem of VASS into linearizability problem of priority queue. Based on above discussion, we have the following complexity result.} %The detailed proof can be found in Appendix \ref{subsec:appendix proof and definition in subsection decidability result}.
%
%%\begin{restatable}{theorem}{ComplexityOfPriorityQueue}
%%\label{theorem:complexity of priority queue}
%%Verifying whether an implementation in $\mathcal{C}$ is linearizable w.r.t. $\seqPQ$ is PSPACE-complete for a fixed number of %threads, and EXPSPACE-complete otherwise.
%%\end{theorem}
